---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ehRsim

<!-- badges: start -->
<!-- badges: end -->

The goal of ehRsim is to provide a set of basic tools for creating simulated electronic health record (EHR) datasets. Tools include generators for continuous, categorical, date and time variables. 

## Installation

You can install the development version of ehRsim from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ssmithm/ehRsim")
```

## Example

This is a basic example developing a few common datasets. Note that this example makes fairly heavy use of [dplyr](https://dplyr.tidyverse.org/):

```{r example_cohort, messages = FALSE, echo = FALSE}
library(ehRsim)
library(dplyr)

#### user input ####
# number of EHR sources
n_src = 7
# number of patients per EHR source
n_p = 1000
# avg number of encounters per patient
aep = 15

# create cohort
cohort <- create_cohort(n_src, n_p)
```


A demographics file can then be created and joined to the cohort table:

```{r example_demographics}
#### demographics ####
sex <- gen_var_char(v = sex, nc = cohort, opts = c("F","M"), p = c(0.53, 0.47))

race <- gen_var_char(v = race,
                     nc = cohort,
                     opts = c("White",
                              "Black/African American",
                              "Asian",
                              "Other",
                              "Multiple Races",
                              NA_character_),
                     p = c(0.55, 0.27, 0.08, 0.03, 0.02, 0.05))

ethnicity <- gen_var_char(v = ethnicity,
                          nc = cohort,
                          opts = c("Hispanic", "Non-Hispanic", NA_character_),
                          p = c(0.18, 0.7, 0.12))

birth_date <- rdate(v = birth_date,
                    nc = cohort,
                    min = eval(Sys.Date() - (365.25*120))) # max age = 120 years

# bind them together for demographics table
demographics <- dplyr::bind_cols(cohort, sex, race, ethnicity, birth_date)
```

You might also want an encounter table, probably with multiple encounters/person:
  
```{r example_encounters}
# first create an encounters shell that has multiple encounters per patient.
encounter_shell <- data.frame(pid = sample(cohort$pid, size = (nrow(cohort) * aep), replace = TRUE)) |>
  dplyr::arrange(pid)

# create an admission date
admit_date <- rdate(v = admit_date,
                    nc = encounter_shell,
                    min = "2012-01-01")

# create a length of stay
los <- data.frame(los = rnbinom(n = nrow(encounter_shell), size = 10, prob = 0.7))

# using the length of stay, create a discharge_date
admit_date <- dplyr::bind_cols(admit_date, los) |>
  dplyr::mutate(discharge_date = admit_date + los)

# admission time
admit_time <- rtime(v = admit_time,
                    nc = encounter_shell,
                    format = "24H",
                    pr.wh = 0.7)

# discharge time
discharge_time <- rtime(v = discharge_time,
                        nc = encounter_shell,
                        format = "24H",
                        pr.wh = 0.6)

# encounter type -- here I use values from the PCORnet common data model. 
enc_type <- gen_var_char(v = enc_type,
                         nc = encounter_shell,
                         opts = c("IP", "AV", "ED", "TH", "OT", "UN"),
                         p = c(0.2, 0.55, 0.08, 0.05, 0.07, 0.05))

# encounter disposition (Alive, Expired, Unknown, Other; again from the PCORnet
# common data model
discharge_disposition <- gen_var_char(v = discharge_disposition,
                                      nc = encounter_shell,
                                      opts = c("A","E","UN","OT"),
                                      p = c(0.8, 0.05, 0.07, 0.08))

# bind encounter table together and 
encounters <- dplyr::bind_cols(dplyr::left_join(encounter_shell, cohort, by = "pid"),
                               admit_date,
                               admit_time,
                               enc_type,
                               discharge_time,
                               discharge_disposition) |>
  dplyr::mutate(discharge_date = dplyr::if_else(enc_type %in% c("AV", "TH", "ED"), admit_date, discharge_date),
                discharge_disposition = dplyr::if_else(enc_type %in% c("AV", "TH"), "A", discharge_disposition))
```


You might have noticed in the above that discharge times are randomly selected. For datasets that have any significant number of observations, it's likely there will be discharge times that come before admission times, even though both admission date and discharge data are equivalent. We can update this with `update_rtime()`:

```{r fix_encounter_datetimes}
encounters_fixed <- encounters |>  
  dplyr::mutate(needs_fixing = ifelse(as.POSIXct(paste(discharge_date, discharge_time), format = "%Y-%m-%d %H:%M") < as.POSIXct(paste(admit_date, admit_time), format = "%Y-%m-%d %H:%M"), "x", "")) 

encounters_fixed |> 
  dplyr::filter(admit_date == discharge_date) |>
  dplyr::select(pid, admit_date, admit_time, discharge_date, discharge_time, needs_fixing) |> 
  head(10)
  
encounters_fixed <- update_rtime(data = encounters_fixed, 
                                 perm_date = admit_date, 
                                 perm_time = admit_time, 
                                 fix_date = discharge_date, 
                                 fix_time = discharge_time)

encounters_fixed |> 
  dplyr::filter(admit_date == discharge_date) |>
  dplyr::select(pid, admit_date, admit_time, discharge_date, discharge_time, needs_fixing) |> 
  head(10) 

encounters_fixed <- encounters_fixed |> 
  dplyr::select(-needs_fixing)
```


We might also want to delete encounters after an Expired discharge_disposition (this will hopefully soon be put into its own function):

```{r clean_up_}
# delete encounters after an Expired discharge_disposition
encounters_death <- encounters_fixed |>
  dplyr::filter(discharge_disposition == "E") |>
  dplyr::select(pid, source, discharge_date) |>
  dplyr::rename(dead_after = discharge_date) |>
  dplyr::arrange(pid, source, dead_after) |>
  dplyr::group_by(source, pid) |>
  dplyr::slice(1) |>
  dplyr::ungroup()

encounters_ <- dplyr::left_join(encounters_fixed, encounters_death, by = c("pid", "source")) |>
  dplyr::arrange(source, pid, admit_date)
encounters <- encounters_ |>
  dplyr::filter(is.na(dead_after) | (admit_date <= dead_after)) |>
  dplyr::select(-dead_after)

# clean up final dataset
encounters <- encounters |>
  dplyr::mutate(unique_combo = paste0(source, "-", pid, "-", admit_date),
                encounterid = openssl::md5(unique_combo)) |>
  dplyr::select(-los, -unique_combo) |>
  dplyr::relocate(pid, encounterid, enc_type, admit_date, discharge_date, discharge_disposition, source)
```


Additional planned functionality include sampling from commonly-employed datasets and ontologies, for example, International Classification of Diseases (ICD) codes, RXNORM or SNOMED codes, LOINC codes, and so on. Stay tuned. 
